---
title: "NaileR interpretive-lift blind benchmark"
format:
  pdf:
    include-in-header:
      - text: |
          \usepackage{fvextra}
          \usepackage{pdflscape}
          \usepackage{tabularx}
          \usepackage{array}
          \usepackage{booktabs}
          \RecustomVerbatimEnvironment{verbatim}{Verbatim}{breaklines=true,breakanywhere=true,fontsize=\small}
execute:
  echo: false
  results: asis
  warning: false
  message: false
---

This report is intentionally rendered only after the blinded responses have
been collected. It contains no condition reveal.

## Method

The benchmark compares two prompt conditions. Every pair was generated with
the same statistical evidence, data, options, introduction, request,
provider, model, and full sampling (`sample.pct = 1` where available).
`generate = FALSE` was used throughout the preparation step.

```{r, results='asis'}
library(knitr)

case_catalog <- data.frame(
  case_id = sprintf("CASE_%02d", 1:7),
  `function` = c("nail_qda", "nail_condes", "nail_condes", "nail_catdes",
                 "nail_catdes", "nail_descfreq", "nail_textual"),
  mode = c("product profile", "standard", "latent", "standard", "latent",
           "description", "grouped textual responses"),
  data = c("SensoMineR::sensochoc", "FactoMineR::decathlon, Points",
           "FactoMineR::decathlon PCA Dim1", "datasets::iris, Species",
           "NaileR::atomic_habit_clust, constructed groups",
           "NaileR::beard_cont", "NaileR::fabric, Fabric == 'A'"),
  selected = c("choc1", "<all>", "<all>", "setosa", "1", "B1", "0"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

case_code <- c(
  "data(chocolates, package = 'SensoMineR'); nail_qda(sensochoc, '~Product+Panelist', firstvar = 5, isolate.groups = TRUE, sample.pct = 1, generate = FALSE)",
  "data(decathlon, package = 'FactoMineR'); nail_condes(decathlon, num.var = 12, interpretation_mode = 'standard', sample.pct = 1, generate = FALSE)",
  "pca <- FactoMineR::PCA(decathlon[, 1:10], scale.unit = TRUE, graph = FALSE); work <- data.frame(Dim1 = pca$ind$coord[, 1], decathlon[, 1:10]); nail_condes(work, num.var = 1, interpretation_mode = 'latent', target_label = 'Dim1', sample.pct = 1, generate = FALSE)",
  "data(iris); nail_catdes(iris, num.var = 5, interpretation_mode = 'standard', isolate.groups = TRUE, quali.sample = 1, quanti.sample = 1, generate = FALSE)",
  "data(atomic_habit_clust, package = 'NaileR'); work <- atomic_habit_clust[, c(1:20, 51)]; nail_catdes(work, num.var = ncol(work), interpretation_mode = 'latent', isolate.groups = TRUE, quali.sample = 1, quanti.sample = 1, generate = FALSE)",
  "data(beard_cont, package = 'NaileR'); nail_descfreq(beard_cont, interpretation_mode = 'description', isolate.groups = TRUE, sample.pct = 1, generate = FALSE)",
  "data(fabric, package = 'NaileR'); work <- droplevels(fabric[fabric$Fabric == 'A', , drop = FALSE]); nail_textual(work, num.var = 4, num.text = 3, isolate.groups = TRUE, sample.pct = 1, seed = 123, generate = FALSE)"
)

case_catalog$code <- case_code
for (i in seq_len(nrow(case_catalog))) {
  cat("- `", case_catalog$case_id[i], "` — ", case_catalog$`function`[i],
      " (", case_catalog$mode[i], "; ", case_catalog$data[i], "; unité: ",
      case_catalog$selected[i], ")\n", sep = "")
}
```

## Evidence verification

The preparation run compared the hashes of the complete canonical evidence and
the selected evidence for each pair. Every pair matched. The latent cases are
also expected to have identical prompts because their prompt branches were not
part of the interpretive-lift change.

```{r}
evidence_check <- data.frame(
  case_id = sprintf("CASE_%02d", 1:7),
  complete_evidence = c(
    "6ef6cb8c60ca485a3c9bcc1b0412f059",
    "6f1cc1d39a055a78145aa2ca3d2dd519",
    "fac0e30ae7e3a9294acab10184128d38",
    "0dee715e69b4919aed1192d70f4b13f5",
    "f4e964ce5ab2df4e8e7521e0a7be2501",
    "8cc93cf8ad746d6b614a13dac99d006d",
    "6440b230c56e124ae167bab5c05489a2"
  ),
  selected_evidence = c(
    "34bc4f42b5bda1302b1654f83fd4369f",
    "6f1cc1d39a055a78145aa2ca3d2dd519",
    "fac0e30ae7e3a9294acab10184128d38",
    "8ffe8c7621222ac3dab47dfe9d5911d7",
    "828613412e67933609c2e30ecb933fb7",
    "04dc257731a4d1c6bc674862c9f7be3f",
    "4208a06ae173b7fcda739090ea0bd300"
  ),
  identical_within_pair = TRUE,
  stringsAsFactors = FALSE
)
kable(evidence_check[, c("case_id", "identical_within_pair")])
```

## Blind prompt and response review

```{r, results='asis'}
read_block <- function(path, blind_id) {
  lines <- readLines(path, warn = FALSE, encoding = "UTF-8")
  begin <- match(paste0("===== ", blind_id, " ====="), lines)
  end <- match(paste0("===== END ", blind_id, " ====="), lines)
  if (is.na(begin) || is.na(end) || end <= begin) {
    stop("Missing block: ", blind_id)
  }
  paste(lines[(begin + 1):(end - 1)], collapse = "\n")
}

block_ids <- function(path, prefix = "CASE_") {
  lines <- readLines(path, warn = FALSE, encoding = "UTF-8")
  starts <- lines[grepl(paste0("^===== ", prefix), lines)]
  gsub("^===== | =====$", "", starts)
}

prompt_ids <- block_ids("prompt_bundle_blinded.md")
response_ids <- block_ids("responses_blinded.md")
stopifnot(
  length(prompt_ids) == 14L,
  length(response_ids) == 14L,
  length(unique(prompt_ids)) == 14L,
  length(unique(response_ids)) == 14L,
  setequal(prompt_ids, response_ids)
)

for (case_id in case_catalog$case_id) {
  case_row <- case_catalog[case_catalog$case_id == case_id, , drop = FALSE]
  evidence_row <- evidence_check[evidence_check$case_id == case_id, , drop = FALSE]
  cat("\n## ", case_id, "\n\n", sep = "")
  cat("**Function:** ", case_row$`function`, "  \n", sep = "")
  cat("**Mode:** ", case_row$mode, "  \n", sep = "")
  cat("**Data:** ", case_row$data, "  \n", sep = "")
  cat("**Selected unit:** ", case_row$selected, "\n\n", sep = "")
  cat("**Minimal R code**\n\n```r\n", case_row$code, "\n```\n\n", sep = "")
  cat("**Evidence check:** complete and selected evidence hashes match within the pair;",
      "the statistical evidence is identical.\n\n", sep = " ")
  for (blind_id in paste0(case_id, c("_X", "_Y"))) {
    cat("### Condition ", substr(blind_id, nchar(blind_id), nchar(blind_id)),
        " (", blind_id, ")\n\n", sep = "")
    cat("#### Prompt exact\n\n```text\n", read_block("prompt_bundle_blinded.md", blind_id),
        "\n```\n\n", sep = "")
    cat("#### Réponse ChatGPT exacte\n\n```text\n", read_block("responses_blinded.md", blind_id),
        "\n```\n\n", sep = "")
    cat("#### Evaluation\n\n")
    cat("| Criterion | Score | Comment |\n|---|---:|---|\n")
    cat("| Interpretive value |  |  |\n")
    cat("| Grounding / traceability |  |  |\n")
    cat("| Calibration |  |  |\n")
    cat("| Synthesis |  |  |\n")
    cat("| Unsupported overreach |  |  |\n")
    cat("| Paraphrase tendency |  |  |\n\n")
    cat("**Qualitative comment:**\n\n\n")
  }
}
```

For the first five criteria use `0 = insufficient`, `1 = acceptable`, and
`2 = convincing`. For Paraphrase tendency use `0 = low`, `1 = moderate`, and
`2 = high`. Add a qualitative comment for every criterion. Do not calculate a
global score automatically.

## Archived blind evaluation

The following section is preserved verbatim from `blind_evaluation.md`. It was
completed before the condition key was consulted; its scores, comments and
blind preferences are not modified by the reveal below.

```{r, results='asis'}
cat(readLines("blind_evaluation.md", warn = FALSE, encoding = "UTF-8"),
    sep = "\n")
```

# Condition reveal and conclusions

The condition key is now consulted. The source-version mapping is displayed
only in this final section.

```{r, results='asis'}
reveal <- data.frame(
  case_id = sprintf("CASE_%02d", 1:7),
  function_name = c("nail_qda", "nail_condes", "nail_condes", "nail_catdes",
                    "nail_catdes", "nail_descfreq", "nail_textual"),
  mode = c("product profile", "standard", "latent", "standard", "latent",
           "description", "grouped textual responses"),
  x_version = c("baseline_2.1.0", "baseline_2.1.0", "interpretive_lift",
                "interpretive_lift", "interpretive_lift", "baseline_2.1.0",
                "interpretive_lift"),
  y_version = c("interpretive_lift", "interpretive_lift", "baseline_2.1.0",
                "baseline_2.1.0", "baseline_2.1.0", "interpretive_lift",
                "baseline_2.1.0"),
  blind_preference = c("Y, avantage léger", "Y, avantage léger",
                       "égalité, témoin inchangé", "X, avantage net",
                       "égalité, témoin inchangé", "Y, avantage léger avec réserve",
                       "X, avantage net"),
  decision = c("KEEP", "KEEP", "UNCHANGED WITNESS", "KEEP", "UNCHANGED WITNESS",
               "ADJUST", "KEEP"),
  comment = c(
    "Le concept sensoriel de niveau supérieur reste explicitement relié aux attributs affichés.",
    "Le gain principal est une meilleure séparation entre synthèse substantive et résultat statistique.",
    "Les prompts latent sont substantivement identiques ; la capacité de nommer le continuum reste le témoin de référence.",
    "La synthèse morphologique apporte un lift interprétatif net sans perte de grounding.",
    "Les prompts latent sont substantivement identiques. Le nom utile mais trop comportemental ‘adopters’ appelle une réserve indépendante.",
    "Le lift est réel, mais le nom ‘Polished Urban Classic’ donne peut-être trop de poids aux signaux rares ‘Parisian bobo’ et ‘educated’.",
    "La synthèse thématique construit une interprétation de niveau supérieur sans inventer de motivations cachées."
  ),
  stringsAsFactors = FALSE
)

for (i in seq_len(nrow(reveal))) {
  cat("## ", reveal$case_id[i], " — ", reveal$function_name[i], " (",
      reveal$mode[i], ")\n\n", sep = "")
  cat("- **Condition X:** `", reveal$x_version[i], "`\n", sep = "")
  cat("- **Condition Y:** `", reveal$y_version[i], "`\n", sep = "")
  cat("- **Préférence aveugle :** ", reveal$blind_preference[i], "\n", sep = "")
  cat("- **Décision :** `", reveal$decision[i], "`\n", sep = "")
  cat("- **Commentaire :** ", reveal$comment[i], "\n\n", sep = "")
}

table_view <- reveal
table_view$comment <- c(
  "Grounded lift.",
  "Clearer calibration.",
  "Latent witness.",
  "Grounded synthesis.",
  "Naming caution.",
  "Sparse-signal caution.",
  "Grounded thematic lift."
)
cat("\\begin{landscape}\n\\scriptsize\n")
cat("\\resizebox{\\linewidth}{!}{%\n")
print(kable(table_view[, c("case_id", "function_name", "mode", "x_version", "y_version",
                           "blind_preference", "decision", "comment")],
            format = "latex", booktabs = TRUE, longtable = FALSE, escape = TRUE,
            col.names = c("Case", "Function", "Mode", "Baseline condition",
                          "Revised condition", "Blind preference", "Decision",
                          "Comment")))
cat("}\n")
cat("\\end{landscape}\n")
```

## Cross-case conclusion

The benchmark supports the hypothesis that the revised formulations increase
interpretive lift without an important general degradation of grounding or
epistemic calibration. The effect is not a generic permission to speculate:

- **Interpretive lift:** the preferred revised responses more often synthesize
  several displayed results into a defensible higher-level concept.
- **Grounding:** the preferred interpretations retain identifiable support in
  the displayed evidence; the latent witnesses remain unchanged.
- **Calibration:** the strongest revised responses distinguish a statistical
  fact from its substantive interpretation and from any broader hypothesis.
- **Unsupported overreach:** no material degradation appears overall, but
  DESCFREQ shows a concrete boundary case in which a concise label can amplify
  sparse contextual signals. CATDES latent shows a separate naming issue: the
  useful label `Broadly capable, low-constraint adopters` can suggest effective
  adoption although the evidence concerns perceived capability and constraint.

The central criterion remains whether the higher-level interpretation is
reasonably defensible from the displayed evidence. A concept absent from the
literal variable names is not automatically overreach when its supporting
pattern is traceable. The appropriate next step is therefore selective
calibration, not a general return to evidence-only paraphrase.
