---
title: "Understanding a qualitative target with nail_catdes()"
author: "NaileR"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Understanding a qualitative target with nail_catdes()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

> **PASS 2 is optional.**
> You do not need `nail_catdes_ground()` to use `nail_catdes()`.
> Use PASS 2 when you want an explicit critical review of how far each interpretation moves beyond the selected evidence.

> **Key principle:** Epistemic temperature tells you how far a statement moves from direct observation; grounding tells you whether the available evidence is sufficient to support that statement as worded.

## From statistical characteristics to interpretation — and, optionally, to epistemic review

This guide presents a practical workflow for using the current `nail_catdes()` family in NaileR.

The workflow has two independent passes:

```text
PASS 1 — nail_catdes()
statistical evidence
        ↓
local semantic interpretation

PASS 2 — nail_catdes_ground()
frozen PASS 1 interpretation
        +
local statistical evidence
        ↓
epistemic review
```

**PASS 2 is optional.** An experienced analyst may stop after `nail_catdes()`.
`nail_catdes_ground()` is useful when you want a rapid critical reading of the interpretation: which statements are direct facts, which are semantic recompositions, which are interpretations, and which are hypotheses that go beyond what was directly measured?

---

## The key idea

`nail_catdes()` does not ask the LLM to calculate the statistics.

R first selects statistically relevant characteristics and rewrites them as explicit factual statements. The LLM then works from those facts to construct a local semantic interpretation for each category or group.

The important distinction is:

```text
R     → statistical evidence
LLM   → semantic interpretation
```

The optional `nail_catdes_ground()` then asks a second analytical pass to review the already-produced interpretation against the local evidence.

```text
PASS 1 interpretation
        ↓
fact / semantic pattern / interpretation / hypothesis
        ↓
supported / contradicted / insufficient / mixed
```

The original PASS 1 response is preserved.

---

# 2. A complete observed-category example

The example below creates three observed tea-consumption segments:

- `Routine`
- `Explorer`
- `Involved`

The names are known in advance. They are therefore **observed categories**, not latent profiles.

```{r, eval=FALSE}
library(NaileR)

set.seed(20260819)

n <- 240

segment <- factor(
  rep(c("Routine", "Explorer", "Involved"), each = n / 3)
)

tea <- data.frame(
  specialist_shop = factor(c(
    sample(c("yes", "no"), n / 3, TRUE, c(0.15, 0.85)),
    sample(c("yes", "no"), n / 3, TRUE, c(0.45, 0.55)),
    sample(c("yes", "no"), n / 3, TRUE, c(0.80, 0.20))
  )),

  loose_leaf = factor(c(
    sample(c("yes", "no"), n / 3, TRUE, c(0.10, 0.90)),
    sample(c("yes", "no"), n / 3, TRUE, c(0.50, 0.50)),
    sample(c("yes", "no"), n / 3, TRUE, c(0.85, 0.15))
  )),

  organic = factor(c(
    sample(c("yes", "no"), n / 3, TRUE, c(0.20, 0.80)),
    sample(c("yes", "no"), n / 3, TRUE, c(0.65, 0.35)),
    sample(c("yes", "no"), n / 3, TRUE, c(0.55, 0.45))
  )),

  consumption_frequency = c(
    rnorm(n / 3, 2.5, 0.7),
    rnorm(n / 3, 4.5, 0.8),
    rnorm(n / 3, 6.5, 0.7)
  ),

  willingness_to_pay = c(
    rnorm(n / 3, 3.5, 0.8),
    rnorm(n / 3, 5.5, 0.9),
    rnorm(n / 3, 7.5, 0.8)
  ),

  Segment = segment
)
```

Here `Segment` is the target variable and is the last column:

```{r, eval=FALSE}
num.var <- ncol(tea)
```

---

# 3. First inspect PASS 1 without calling an LLM

A useful habit is to inspect what NaileR intends to send to the model before generating anything.

```{r, eval=FALSE}
pass1_preview <- nail_catdes(
  tea,
  num.var = num.var,

  introduction = paste(
    "A study investigates tea purchasing and consumption practices.",
    "The three segments are observed categories defined before this analysis."
  ),

  request = paste(
    "Explain the profile of each segment by combining the selected facts.",
    "Keep the original segment names."
  ),

  interpretation_mode = "standard",
  quali.sample = 0.10,
  quanti.sample = 0.10,
  drop.negative = FALSE,
  generate = FALSE
)

cat(pass1_preview)
```

## What should you look at?

First, verify that the **introduction** correctly describes the analytical situation. It is not decorative: it tells the LLM what kind of object it is interpreting.

Second, verify the **request**. It should express the actual analytical question rather than merely asking for a generic summary.

Third, inspect the `Data` section. The statements should be understandable without having to decode a statistical table. For example:

```text
"yes" is MORE FREQUENT in this group than in the full sample
"no" is LESS FREQUENT in this group than in the full sample
the mean is HIGHER in this group than in the full sample
```

These directions must be read literally.

Finally, check that the prompt is local: the facts shown for one group should concern that group only.

---

# 4. Inspect the statistical facts behind PASS 1

For a more detailed audit:

```{r, eval=FALSE}
semantic_facts <- attr(
  pass1_preview,
  "semantic_facing_evidence"
)

semantic_facts$groups$Involved$text
```

A typical pattern for `Involved` is:

```text
loose leaf:
  "no"  LESS FREQUENT
  "yes" MORE FREQUENT

consumption frequency:
  HIGHER
```

This is the factual basis of the interpretation.

For binary qualitative variables, NaileR may display both sides of the significant contrast. This makes polarity easier to read and reduces errors such as interpreting "`no` is less frequent" in the wrong direction.

---

# 5. Generate PASS 1

Choose the LLM backend and model that you want to use.

Example with Ollama:

```{r, eval=FALSE}
pass1 <- nail_catdes(
  tea,
  num.var = num.var,

  introduction = paste(
    "A study investigates tea purchasing and consumption practices.",
    "The three segments are observed categories defined before this analysis."
  ),

  request = paste(
    "Explain the profile of each segment by combining the selected facts.",
    "Keep the original segment names."
  ),

  interpretation_mode = "standard",
  quali.sample = 0.10,
  quanti.sample = 0.10,
  drop.negative = FALSE,

  provider = "ollama",
  model = "mistral-small3.2",
  generate = TRUE
)
```

Display the combined interpretation:

```{r, eval=FALSE}
cat(pass1$response)
```

Or inspect the local response for one category:

```{r, eval=FALSE}
profiles <- attr(pass1, "semantic_profiles")

cat(
  profiles$groups$Involved$response
)
```

## What should you look at in PASS 1?

Do not only ask whether the response sounds plausible.

Read it in this order:

1. **Does it correctly reproduce the statistical direction?**
2. **Does it combine several facts into a meaningful pattern?**
3. **Does it introduce concepts that were not directly measured?**
4. **Are those additional concepts useful interpretations, or are they being stated too strongly?**

For example, from:

```text
more loose-leaf tea
+
higher tea-consumption frequency
```

an interpretation such as:

```text
greater involvement with tea
```

may be useful.

But statements such as:

```text
greater expertise
values tradition
values quality
```

go further than the variables directly establish.

That does not make them useless. It means they should be read as hypotheses rather than as direct results.

---

# 6. PASS 2 is optional

If you are comfortable making this distinction yourself, you can stop here.

```{r, eval=FALSE}
pass1
```

`nail_catdes_ground()` is intended for users who want a rapid, explicit critical review of the PASS 1 interpretation.

It does **not** regenerate or rewrite PASS 1.

---

# 7. Preview the epistemic review

Before calling the reviewer:

```{r, eval=FALSE}
ground_preview <- nail_catdes_ground(
  pass1,
  generate = FALSE
)
```

Inspect one grounding prompt:

```{r, eval=FALSE}
cat(
  ground_preview$grounding_prompts$Explorer
)
```

The prompt contains:

```text
Frozen PASS 1 Response
+
Local Evidence
+
Epistemic Contract
```

The evidence references shown to the LLM (`G001E001`, `G001E002`, ...) are short local references. NaileR maps them back to the canonical internal Evidence IDs after generation.

---

# 8. Generate PASS 2

```{r, eval=FALSE}
ground <- nail_catdes_ground(
  pass1,
  generate = TRUE
)
```

By default, `nail_catdes_ground()` inherits the model and provider recorded in PASS 1. You can override them explicitly if required.

PASS 2 performs one independent local review per eligible group.

---

# 9. Read one grounded profile

A small display helper is convenient:

```{r, eval=FALSE}
print_ground_profile <- function(x, group) {

  profile <- x$grounded_profiles[[group]]

  cat("\n##", group, "\n\n")

  for (a in profile$assertions) {

    cat(
      sprintf(
        "[%s | temperature=%d | %s | support=%s]\n",
        toupper(a$epistemic_level),
        a$epistemic_temperature,
        toupper(a$grounding$status),
        toupper(a$grounding$support_type)
      )
    )

    cat(a$text, "\n")

    if (length(a$grounding$supporting_evidence_ids)) {
      cat(
        "supporting evidence:",
        paste(
          a$grounding$supporting_evidence_ids,
          collapse = " | "
        ),
        "\n"
      )
    }

    if (length(a$grounding$contradicting_evidence_ids)) {
      cat(
        "contradicting evidence:",
        paste(
          a$grounding$contradicting_evidence_ids,
          collapse = " | "
        ),
        "\n"
      )
    }

    cat("rationale:", a$grounding$rationale, "\n\n")
  }
}
```

Then:

```{r, eval=FALSE}
print_ground_profile(
  ground,
  "Explorer"
)

print_ground_profile(
  ground,
  "Involved"
)

print_ground_profile(
  ground,
  "Routine"
)
```

---

# 10. How to read epistemic temperature

Temperature represents the **distance from direct observation**, not truth.

| Temperature | Level | Meaning |
|---:|---|---|
| 0 | `fact` | Directly restatable from the statistical evidence |
| 1 | `semantic_pattern` | Recomposition of convergent facts |
| 2 | `interpretation` | Higher-level meaning given to the pattern |
| 3 | `hypothesis` | Explanation, extension, mechanism or unmeasured characteristic |

A hot statement is not necessarily bad.

A hypothesis may be scientifically useful. The important point is to know that it is a hypothesis.

---

# 11. Grounding is a different question

Grounding asks whether the local evidence is sufficient for the assertion **as worded**.

| Status | Meaning |
|---|---|
| `supported` | The listed evidence is sufficient to justify the assertion |
| `contradicted` | The local evidence points against the assertion |
| `insufficient` | Evidence may be relevant, but does not establish the assertion |
| `mixed` | Evidence genuinely points in more than one direction |

Temperature and grounding are therefore independent.

For example:

```text
FACT + SUPPORTED
"The mean consumption frequency is higher."

HYPOTHESIS + INSUFFICIENT
"The group may be more knowledgeable about tea."

INTERPRETATION + INSUFFICIENT
"The observed organic purchasing pattern reflects environmental values."
```

A particularly important distinction is:

```text
compatible with the evidence
≠
supported by sufficient evidence
```

---

# 12. Understand the support type

Each reviewed assertion also receives a support type:

| Support type | Meaning |
|---|---|
| `direct` | The evidence nearly restates the assertion |
| `convergent` | Several pieces of evidence jointly support the assertion |
| `indirect` | The evidence is relevant but the assertion passes through an interpretation or proxy |
| `none` | No listed evidence genuinely supports the assertion |

A useful combination is:

```text
INSUFFICIENT + INDIRECT
```

This means:

> there is a real connection to the data, but the data are not enough to establish the statement.

This is often the appropriate status for a useful analytical hypothesis.

---

# 13. A practical reading order

When reviewing a grounded profile, use the following order.

### First: cold, directly supported statements

Look for:

```text
FACT
temperature = 0
SUPPORTED
DIRECT
```

These are the statistical anchors of the profile.

### Second: convergent semantic patterns

Look for:

```text
SEMANTIC_PATTERN
temperature = 1
SUPPORTED
CONVERGENT
```

These are often the most useful bridge between statistical description and substantive interpretation.

### Third: interpretations

Interpretations may be supported, or they may be only indirectly grounded.

Ask:

> Is this meaning genuinely supported by several facts, or is it simply plausible?

### Fourth: hypotheses

Look especially for:

```text
HYPOTHESIS
temperature = 3
INSUFFICIENT
INDIRECT
```

Do not discard them. Treat them as questions for further investigation.

### Finally: contradicted statements

Any:

```text
CONTRADICTED
```

deserves immediate attention.

This may reveal a polarity inversion, an incorrect reading of a modality, or an interpretation that conflicts with the actual local evidence.

---

# 14. Quickly inspect only the assertions that need caution

A useful helper:

```{r, eval=FALSE}
collect_assertions <- function(x) {

  do.call(
    rbind,
    lapply(
      names(x$grounded_profiles),
      function(group) {

        assertions <-
          x$grounded_profiles[[group]]$assertions

        if (!length(assertions))
          return(NULL)

        do.call(
          rbind,
          lapply(
            assertions,
            function(a) {
              data.frame(
                group = group,
                text = a$text,
                epistemic_level =
                  a$epistemic_level,
                temperature =
                  a$epistemic_temperature,
                grounding =
                  a$grounding$status,
                support =
                  a$grounding$support_type,
                stringsAsFactors = FALSE
              )
            }
          )
        )
      }
    )
  )
}

review_table <- collect_assertions(ground)
```

Show only assertions that are not fully supported:

```{r, eval=FALSE}
subset(
  review_table,
  grounding != "supported"
)
```

Show only hypotheses:

```{r, eval=FALSE}
subset(
  review_table,
  epistemic_level == "hypothesis"
)
```

Show the hottest assertions first:

```{r, eval=FALSE}
review_table[
  order(
    review_table$temperature,
    decreasing = TRUE
  ),
]
```

This is often the fastest way to use PASS 2 in practice.

---

# 15. Inspect computation time

`nail_catdes_ground()` records wall-clock and CPU timing.

Total PASS 2 time:

```{r, eval=FALSE}
ground$timing$total
```

For one group:

```{r, eval=FALSE}
ground$timing$grounding_calls$Explorer
```

This separates:

```text
total
backend
parse_and_validate
```

For local LLMs, the wall-clock `elapsed_seconds` is generally the most useful quantity.

A convenient summary:

```{r, eval=FALSE}
data.frame(
  group = names(ground$timing$grounding_calls),

  elapsed = vapply(
    ground$timing$grounding_calls,
    function(x)
      x$total$elapsed_seconds,
    numeric(1)
  ),

  backend = vapply(
    ground$timing$grounding_calls,
    function(x)
      x$backend$elapsed_seconds,
    numeric(1)
  ),

  parse_validate = vapply(
    ground$timing$grounding_calls,
    function(x)
      x$parse_and_validate$elapsed_seconds,
    numeric(1)
  )
)
```

The distinction is useful because LLM inference time may dominate the R processing time.

---

# 16. Observed categories versus latent groups

The choice of `interpretation_mode` is important.

## Observed categories

Use:

```{r, eval=FALSE}
interpretation_mode = "standard"
```

when categories already have a substantive meaning:

```text
Species
Country
Product
Treatment
Segment defined before the analysis
```

NaileR preserves their names and interprets what statistically characterizes each category.

## Constructed or latent groups

Use:

```{r, eval=FALSE}
interpretation_mode = "latent"
```

when the groups were constructed by an analytical method, for example clustering.

In this case, the group identifiers are not substantive meanings. The LLM may reconstruct the common semantic pattern and propose an interpretive name.

Example:

```{r, eval=FALSE}
library(FactoMineR)

data(local_food, package = "NaileR")

set.seed(1)

food_mca <- MCA(
  local_food,
  quali.sup = 46:63,
  ncp = 30,
  level.ventil = 0.05,
  graph = FALSE
)

food_hcpc <- HCPC(
  food_mca,
  nb.clust = 3,
  graph = FALSE
)

food_cluster_data <- food_hcpc$data.clust

latent_pass1 <- nail_catdes(
  food_cluster_data,
  num.var = ncol(food_cluster_data),

  introduction = paste(
    "A study on sustainable food systems was conducted with French participants.",
    "The groups were obtained by clustering individuals in an MCA factorial space."
  ),

  request = paste(
    "For each group, infer the common meaning of the strongest convergent characteristics.",
    "Explain what the group seems to represent and propose a concise interpretive name."
  ),

  interpretation_mode = "latent",
  quali.sample = 0.10,
  quanti.sample = 0.10,
  drop.negative = FALSE,

  provider = "ollama",
  model = "mistral-small3.2",
  generate = TRUE
)

latent_ground <- nail_catdes_ground(
  latent_pass1,
  generate = TRUE
)
```

In a latent analysis, pay particular attention to broad interpretive labels. A proposed group name is necessarily a semantic abstraction and should be read in light of the underlying assertions.

---

# 17. Parameters worth thinking about

## `introduction`

Describe the study, population, origin of the target variable and any context required to interpret the evidence.

A good introduction reduces ambiguity.

## `request`

State the substantive analytical objective.

Compare:

```text
"Describe the groups."
```

with:

```text
"Explain how the groups differ in their relationship to local food policy,
identify the strongest convergent characteristics, and avoid causal claims."
```

The second request gives the model a much clearer analytical task.

## `interpretation_mode`

Use:

```{r, eval=FALSE}
"standard"
```

for meaningful observed categories, and:

```{r, eval=FALSE}
"latent"
```

for constructed profiles/classes.

## `quali.sample` and `quanti.sample`

These control how much selected evidence enters the semantic interpretation.

A smaller value produces a more selective profile. A larger value provides a richer but potentially more crowded prompt.

There is no universally optimal value. Start with a selective specification and inspect the semantic-facing evidence.

## `drop.negative`

Keeping:

```{r, eval=FALSE}
drop.negative = FALSE
```

is often useful.

Under-representation and lower-than-average characteristics may be essential for understanding a group. A negative statistical direction is not a negative substantive judgment.

---

# 18. Advanced audit objects

Most users do not need these objects for ordinary interpretation, but they make the workflow auditable.

From PASS 1:

```{r, eval=FALSE}
attr(pass1, "statistical_profiles")
attr(pass1, "interpretation_evidence")
attr(pass1, "semantic_facing_evidence")
attr(pass1, "local_prompts")
attr(pass1, "semantic_profiles")
attr(pass1, "catdes_settings")
```

From PASS 2:

```{r, eval=FALSE}
ground$semantic_profiles
ground$grounding_prompts
ground$evidence_reference_maps
ground$grounded_profiles
ground$settings
ground$metadata
ground$timing
```

The important architectural property is that the PASS 1 semantic response is preserved in PASS 2:

```{r, eval=FALSE}
ground$grounded_profiles$Explorer$raw_response
```

PASS 2 reviews PASS 1; it does not silently replace it.

---

# 19. What PASS 2 does not mean

Do **not** interpret:

```text
INSUFFICIENT
```

as:

```text
false
```

It means that the available local evidence does not establish the assertion as worded.

Do **not** interpret:

```text
HYPOTHESIS
```

as:

```text
bad interpretation
```

A hypothesis can be the most scientifically interesting output.

Do **not** interpret:

```text
SUPPORTED
```

as proof of a causal mechanism.

Grounding is relative to the selected local evidence and to the assertion as stated.

Finally, PASS 2 itself is an LLM-assisted critical reading. Its classifications remain analytical judgments that can be inspected and challenged by the user.

---

# 20. Recommended workflow

For routine use:

```{r, eval=FALSE}
# 1. Inspect the PASS 1 prompt
preview <- nail_catdes(..., generate = FALSE)

# 2. Generate local semantic profiles
pass1 <- nail_catdes(..., generate = TRUE)

# 3. Read PASS 1 yourself
cat(pass1$response)

# 4. Optional: inspect PASS 2 before generation
ground_preview <- nail_catdes_ground(
  pass1,
  generate = FALSE
)

# 5. Optional: perform epistemic review
ground <- nail_catdes_ground(
  pass1,
  generate = TRUE
)

# 6. Focus on hot / insufficient / contradicted assertions
review_table <- collect_assertions(ground)

subset(
  review_table,
  temperature >= 2 |
    grounding != "supported"
)
```

The intended analytical progression is:

```text
What statistically characterizes this group?
        ↓
What common meaning emerges from these characteristics?
        ↓
How far does each interpretation move beyond the observed evidence?
        ↓
What should I retain as a result,
what should I phrase cautiously,
and what should I investigate as a hypothesis?
```

That is the practical role of `nail_catdes()` followed, when useful, by `nail_catdes_ground()`.