---
title: "NaileR 2.0.0: A practical guide to the main functions"
author: "Sébastien Lê"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NaileR 2.0.0: A practical guide to the main functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

`NaileR` combines statistical summaries with large language models (LLMs). Its
main functions first transform statistical or textual results into explicit,
auditable prompts. Generation is optional.

A useful development and teaching pattern is therefore:

1. call a function with `generate = FALSE`;
2. inspect the prompt and the attached statistical result;
3. only then set `generate = TRUE` and choose an LLM provider.

```{r}
library(NaileR)
```

# Choosing a function

| Data or objective | Main function | Statistical basis |
|---|---|---|
| Describe a categorical variable | `nail_catdes()` | `FactoMineR::catdes()` |
| Describe a continuous variable | `nail_condes()` | `FactoMineR::condes()` |
| Interpret rows of a contingency table | `nail_descfreq()` | `FactoMineR::descfreq()` |
| Interpret grouped open-ended texts | `nail_textual()` | Grouped raw texts |
| Prepare structured textual profiles | `nail_textual_prep()` | LLM-structured summaries |
| Prepare structured statistical profiles | `nail_group_profile_prep()` | Group characterization |
| Cross statistical and textual profiles | `nail_textual_contextualized()` | Prepared profile objects |
| Interpret QDA product profiles | `nail_qda()` | `SensoMineR::decat()` |
| Prepare product summaries for a sensory space | `nail_qda_spaceprep()` | QDA characterization |
| Interpret dimensions of a sensory product space | `nail_qda_space()` | PCA of adjusted means |
| Sort textual descriptions by similarity | `nail_sort()` | LLM-assisted grouping |

# Prompt-only and generated modes

All major interpretation functions use the same principle.

```{r}
prompt <- nail_catdes(
  dataset = iris,
  num.var = 5,
  generate = FALSE
)

cat(prompt)
```

With `generate = FALSE`, no Ollama or Gemini request is made. The underlying
statistical object remains available as an attribute.

```{r}
catdes_result <- attr(prompt, "catdes_result", exact = TRUE)
names(catdes_result)
```

To generate a response locally with Ollama:

```{r}
result <- nail_catdes(
  dataset = iris,
  num.var = 5,
  provider = "ollama",
  model = "llama3",
  generate = TRUE
)
```

To use Gemini, store the API key in the environment rather than in source code.

```{r}
Sys.setenv(GEMINI_API_KEY = "your-key")

result <- nail_catdes(
  dataset = iris,
  num.var = 5,
  provider = "gemini",
  model = "gemini-2.5-flash",
  generate = TRUE
)
```

# Categorical variables with `nail_catdes()`

Use `nail_catdes()` when the target variable is categorical. The variable may
be an explicit observed category or a class produced by clustering.

```{r}
prompt <- nail_catdes(
  dataset = iris,
  num.var = 5,
  interpretation_mode = "standard",
  prompt_style = "compact",
  generate = FALSE
)
```

For latent or constructed groups, use `interpretation_mode = "latent"`.

```{r}
prompt <- nail_catdes(
  dataset = clustered_data,
  num.var = ncol(clustered_data),
  interpretation_mode = "latent",
  isolate.groups = TRUE,
  drop.negative = FALSE,
  generate = FALSE
)
```

When `isolate.groups = TRUE`, the result is a named list containing one prompt
per group.

# Continuous variables with `nail_condes()`

Use `nail_condes()` when the target is numeric. The variable-level results are
the main evidence; discretized low- and high-end profiles provide complementary
illustrations.

```{r}
prompt <- nail_condes(
  dataset = iris,
  num.var = 1,
  interpretation_mode = "standard",
  target_label = "Sepal length",
  generate = FALSE
)
```

For a score interpreted as a latent continuum:

```{r}
prompt <- nail_condes(
  dataset = score_data,
  num.var = 1,
  interpretation_mode = "latent",
  generate = FALSE
)
```

The original and profile-level `condes` results are stored as attributes.

```{r}
attr(prompt, "condes_result", exact = TRUE)
attr(prompt, "condes_profile_result", exact = TRUE)
```

# Contingency tables with `nail_descfreq()`

`nail_descfreq()` interprets each row through over-represented and
under-represented column attributes.

```{r}
data(beard_cont, package = "NaileR")

prompt <- nail_descfreq(
  dataset = beard_cont,
  interpretation_mode = "description",
  explicit_row_labels = TRUE,
  generate = FALSE
)
```

Use `interpretation_mode = "comparison"` when the main objective is to compare
rows. Set `rows_are_ordered = TRUE` only when the order has substantive meaning.

```{r}
prompt <- nail_descfreq(
  dataset = ordered_table,
  interpretation_mode = "comparison",
  rows_are_ordered = TRUE,
  generate = FALSE
)
```

# Raw textual data with `nail_textual()`

The input must contain a grouping variable and a text variable.

```{r}
text_data <- data.frame(
  group = factor(c("A", "A", "B", "B")),
  comment = c(
    "Fresh, light and easy to eat.",
    "Simple, floral and pleasant.",
    "Dark, intense and slightly bitter.",
    "Strong cocoa character and long persistence."
  )
)

prompts <- nail_textual(
  dataset = text_data,
  num.var = 1,
  num.text = 2,
  isolate.groups = TRUE,
  text_role = "comments",
  generate = FALSE
)
```

The prompts contain raw textual evidence. Sampling can be made reproducible
with `sample.pct` and `seed`.

```{r}
prompts <- nail_textual(
  dataset = text_data,
  num.var = 1,
  num.text = 2,
  sample.pct = 0.75,
  seed = 123,
  generate = FALSE
)
```

# Structured textual and statistical preparation

`nail_textual_prep()` converts each group's texts into a short structured
profile. Because the structured fields are produced by an LLM, a complete
preparation object requires `generate = TRUE`.

```{r}
textual_prep <- nail_textual_prep(
  dataset = text_data,
  num.var = 1,
  num.text = 2,
  provider = "ollama",
  model = "llama3",
  generate = TRUE
)

names(textual_prep[[1]])
```

`nail_group_profile_prep()` performs the corresponding preparation from
statistical group descriptors.

```{r}
profile_prep <- nail_group_profile_prep(
  dataset = analysis_data,
  num.var = group_column,
  provider = "ollama",
  model = "llama3",
  generate = TRUE
)
```

Each generated preparation contains a `parsed` component designed for reuse.

```{r}
textual_prep[[1]]$parsed
profile_prep[[1]]$parsed
```

# Contextualized textual interpretation

`nail_textual_contextualized()` crosses the prepared statistical profile and
the prepared textual profile for common groups.

```{r}
contextualized_prompts <- nail_textual_contextualized(
  group_profile_prep = profile_prep,
  textual_prep = textual_prep,
  interpretation_mode = "comparative",
  include_verbatims = TRUE,
  generate = FALSE
)
```

In no-generation mode, both preparation objects must already exist. This
ensures that `generate = FALSE` remains a strictly offline prompt-building
mode.

To produce the final interpretation:

```{r}
contextualized_result <- nail_textual_contextualized(
  group_profile_prep = profile_prep,
  textual_prep = textual_prep,
  interpretation_mode = "comparative",
  provider = "ollama",
  model = "llama3",
  generate = TRUE
)
```

# QDA and sensory workflows

`nail_qda()` interprets product profiles obtained from a QDA design. The
formula and sensory-variable indices follow the conventions of
`SensoMineR::decat()`.

```{r}
qda_prompt <- nail_qda(
  dataset = sensory_data,
  formul = sensory_attribute ~ product + panelist,
  firstvar = first_sensory_column,
  lastvar = last_sensory_column,
  product_knowledge = "known",
  generate = FALSE
)
```

Use `nail_qda_spaceprep()` to create structured product summaries before
interpreting a multidimensional product space.

```{r}
product_prep <- nail_qda_spaceprep(
  dataset = sensory_data,
  formul = sensory_attribute ~ product + panelist,
  firstvar = first_sensory_column,
  lastvar = last_sensory_column,
  provider = "ollama",
  model = "llama3",
  generate = TRUE
)
```

`nail_qda_space()` performs a PCA of adjusted product means and prepares one
prompt per retained dimension.

```{r}
space_prompts <- nail_qda_space(
  x = product_prep,
  ncp = 3,
  min_inertia_pct = 10,
  expertise_mode = "sensory",
  generate = FALSE
)
```

The multidimensional object is available through the `qda_space` attribute.

```{r}
space <- attr(space_prompts, "qda_space", exact = TRUE)
space$eig
space$retained_axes
```

# Additional textual utilities

`nail_sort()` groups textual descriptions for each assessor or column. Unlike
the prompt-only interpretation functions, the useful sorted result requires
LLM generation.

```{r}
data(beard_wide, package = "NaileR")

sorted <- nail_sort(
  dataset = beard_wide[, 1:5],
  stimulus_id = "beard",
  nb.clusters = 4,
  provider = "ollama",
  model = "llama3",
  generate = TRUE
)
```

The similarity and distance helpers `sim_llm()`, `dist_mat_llm()` and
`dist_ref_llm()` also require an LLM. They are intended for controlled
interactive analyses rather than package checks or reproducible offline
pipelines.

# Recommended workflow

For reproducible analyses:

1. keep statistical preprocessing in R;
2. start with `generate = FALSE` and archive the prompt;
3. inspect the statistical attributes attached to the prompt;
4. record the provider, model and generation options;
5. keep generated interpretations together with their prompts;
6. avoid placing API keys directly in scripts;
7. distinguish statistical evidence from the LLM's narrative synthesis.

# Further help

Function-specific documentation is available through:

```{r}
help(package = "NaileR")
?nail_catdes
?nail_condes
?nail_descfreq
?nail_textual
?nail_qda
```
