---
title: "Evidence-first interpretation with NaileR"
author: "Sébastien Lê"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evidence-first interpretation with NaileR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Evidence first, interpretation second

`NaileR` connects statistical analyses in R with optional interpretation by a
large language model (LLM). Its central rule is:

```text
statistical or textual evidence produced in R
                    ↓
          selected factual evidence
                    ↓
              LLM interpretation
```

Depending on the method, R calculates or collects the results used for
interpretation: for example adjusted means, p-values, statistical tables, or
original text responses. NaileR then selects and renders a method-specific
view of that evidence for a prompt. The LLM may
synthesize and contextualize the displayed facts, but its response is not a
new statistical result. Important interpretations therefore remain subject to
human review, especially when they concern causality, motivations, minority
positions, or constructed latent groups.

The examples below use the historical NaileR and FactoMineR datasets. They are
preview examples and do not require a running LLM server.

```{r packages}
library(NaileR)
library(FactoMineR)
library(SensoMineR)
```

# Inspect before generating

Every rebuilt workflow exposes the same inspection grammar:

```r
result <- nail_xxx(..., generate = FALSE)

nail_evidence(result, select = ...)
nail_prompt(result, select = ..., print = FALSE)

# Available only after generate = TRUE:
nail_response(result, select = ..., print = FALSE)
```

`nail_evidence()` lets you inspect the statistical results or original text
used for interpretation before the LLM generates a response. `nail_prompt()`
exposes the exact rendered prompt; and `nail_response()` exposes the raw LLM
response. Changing `introduction` or `request` changes how the question is
framed for the LLM, without altering the underlying statistical results.

Quantitative Descriptive Analysis (QDA) is a sensory profiling method used to
describe products through evaluated sensory attributes. The following QDA
preview uses a study context and a simple analytical request without
contacting an LLM.

```{r qda-preview}
data(chocolates, package = "SensoMineR")

qda_preview <- nail_qda(
  dataset = sensochoc,
  formul = "~Product+Panelist",
  firstvar = 5,
  introduction = paste(
    "Six chocolates were evaluated by a trained sensory panel.",
    "The objective is to understand their sensory profiles."
  ),
  request = "Describe the overall sensory profile of each chocolate by grouping related attributes together.",
  isolate.groups = TRUE,
  sample.pct = 1,
  generate = FALSE
)

qda_product <- names(nail_evidence(qda_preview)$products)[[1L]]
qda_product_evidence <- nail_evidence(qda_preview, select = qda_product)
qda_product_prompt <- nail_prompt(qda_preview, select = qda_product, print = FALSE)

qda_product_evidence$metrics
cat(substr(qda_product_prompt, 1, 900), "\n")
```

The evidence remains inspectable even when no response exists. To generate a
response, add `provider`, `model`, and `generate = TRUE` in a controlled
environment:

```{r qda-generation, eval = FALSE}
qda_generated <- nail_qda(
  dataset = sensochoc,
  formul = "~Product+Panelist",
  firstvar = 5,
  isolate.groups = TRUE,
  provider = "ollama",
  model = "mistral-small3.2",
  generate = TRUE
)

nail_response(qda_generated, select = "choc1", print = FALSE)
```

# Why combine statistics and an LLM?

Statistical procedures are well suited to calculating contrasts, profiles,
associations, frequencies, coordinates, or textual registries. They do not by
themselves provide every domain-specific label or substantive synthesis an
analyst may need. R provides the statistical results, while the LLM helps
explain their meaning in plain language and in the context of the study.
This remains grounded in the displayed evidence, the user request, and the
interpretive status.

The `introduction` gives study context. It is not statistical evidence. The
`request` states the analytical question. It should say what to interpret and
what distinctions matter. Guardrails in the method-specific reading guide
explain how to read the evidence; they do not turn contextual information into
measurements.

```{r context-request-invariance}
qda_default <- nail_qda(
  dataset = sensochoc,
  formul = "~Product+Panelist",
  firstvar = 5,
  isolate.groups = TRUE,
  generate = FALSE
)
qda_custom <- nail_qda(
  dataset = sensochoc,
  formul = "~Product+Panelist",
  firstvar = 5,
  isolate.groups = TRUE,
  introduction = "A sensory study of chocolate products.",
  request = "Describe the overall sensory profile of each chocolate by grouping related attributes together.",
  generate = FALSE
)

identical(nail_evidence(qda_default), nail_evidence(qda_custom))
```

# Main statistical workflows

## Continuous targets: `nail_condes()`

`nail_condes()` characterizes a continuous target from categorical and
quantitative variables. In `standard` mode the target is an observed variable
whose meaning is already known and whose name is preserved, such as an
observed performance score. In `latent` mode the target may be a constructed
score, component, or dimension whose numerical label is not immediately
meaningful; for example, a PCA axis such as `Dim.1` may need to be interpreted
from the variables that characterize it. The prompt asks the LLM to
reconstruct and name that substantive meaning from the evidence.

### Observed target

```{r condes-standard}
data(decathlon, package = "FactoMineR")

condes_standard <- nail_condes(
  dataset = decathlon,
  num.var = 12,
  interpretation_mode = "standard",
  introduction = "Decathlon performance is described by the observed score.",
  request = "Identify the coherent performance profile associated with the target.",
  generate = FALSE
)

nail_evidence(condes_standard)
cat(substr(nail_prompt(condes_standard, print = FALSE), 1, 900), "\n")
```

### Constructed dimension

```{r condes-latent}
decathlon_active <- decathlon[, 1:10, drop = FALSE]
pca_decathlon <- PCA(decathlon_active, graph = FALSE)
decathlon_dim1 <- data.frame(
  Dim1 = pca_decathlon$ind$coord[, 1],
  decathlon_active,
  check.names = FALSE
)

condes_latent <- nail_condes(
  dataset = decathlon_dim1,
  num.var = 1,
  interpretation_mode = "latent",
  target_label = "Dim1",
  request = "Propose a substantive name for this constructed continuum.",
  generate = FALSE
)

nail_evidence(condes_latent)
cat(substr(nail_prompt(condes_latent, print = FALSE), 1, 900), "\n")
```

## Categorical groups: `nail_catdes_prep()` and `nail_catdes()`

`nail_catdes_prep()` is the mechanical preparation layer. `nail_catdes()`
uses its canonical profiles to build a prompt. In `standard` mode, the
categories are observed and their names are preserved, for example species or
a predefined consumer segment. In `latent` mode, labels identify constructed
groups whose meaning is not immediately clear; the prompt asks for an
interpretive name based on each group's profile.

### Observed categories with `iris`

```{r catdes-standard}
data(iris)

catdes_standard <- nail_catdes(
  dataset = iris,
  num.var = 5,
  interpretation_mode = "standard",
  isolate.groups = TRUE,
  request = "Describe what characterizes each observed species from the displayed evidence.",
  generate = FALSE
)

catdes_profiles <- nail_catdes_prep(
  x = attr(catdes_standard, "statistical_profiles", exact = TRUE)
)
catdes_group <- names(nail_evidence(catdes_standard)$groups)[[1L]]
catdes_profiles$metadata$schema
nail_evidence(catdes_standard, select = catdes_group)
cat(nail_prompt(catdes_standard, select = catdes_group, print = FALSE), "\n")
```

### Constructed groups with `atomic_habit_clust`

```{r catdes-latent}
data(atomic_habit_clust, package = "NaileR")
atomic_habit_profiles <- atomic_habit_clust[, c(1:20, 51), drop = FALSE]

catdes_latent <- nail_catdes(
  dataset = atomic_habit_profiles,
  num.var = ncol(atomic_habit_profiles),
  interpretation_mode = "latent",
  isolate.groups = TRUE,
  request = "Interpret each constructed group from its coherent displayed profile and propose a descriptive name.",
  generate = FALSE
)

latent_group <- names(nail_evidence(catdes_latent)$groups)[[1L]]
nail_evidence(catdes_latent, select = latent_group)
cat(substr(nail_prompt(catdes_latent, select = latent_group, print = FALSE), 1, 1000), "\n")
```

## Contingency profiles: `nail_descfreq()`

`nail_descfreq()` describes rows of a frequency/contingency table. Relative
frequencies are evidence about a profile; they are not causal explanations or
absolute importance scores. The complete frequency evidence remains separate
from the subset selected for the prompt.

```{r descfreq}
data(beard_cont, package = "NaileR")

descfreq_preview <- nail_descfreq(
  dataset = beard_cont,
  interpretation_mode = "description",
  isolate.groups = TRUE,
  sample.pct = 1,
  request = "Characterize each row from the convergent displayed frequency pattern.",
  generate = FALSE
)

descfreq_row <- names(nail_evidence(descfreq_preview)$rows)[[1L]]
nail_evidence(descfreq_preview, select = descfreq_row)
cat(substr(nail_prompt(descfreq_preview, select = descfreq_row, print = FALSE), 1, 1000), "\n")
```

## Grouped open-ended text: `nail_textual_prep()` and `nail_textual()`

`nail_textual()` registers every non-empty text with a stable `TXT######`
identifier. Sampling affects only the texts shown to the LLM; it does not
change the canonical textual registry. `nail_textual_prep()` is the historical
structured preparation workflow and remains available.

```{r textual}
data(fabric, package = "NaileR")
fabric_A <- droplevels(fabric[fabric$Fabric == "A", , drop = FALSE])

textual_preview <- nail_textual(
  dataset = fabric_A,
  num.var = 4,
  num.text = 3,
  isolate.groups = TRUE,
  sample.pct = 0.35,
  seed = 123,
  text_role = "responses",
  request = paste(
    "Identify recurring frames while preserving internal diversity.",
    "Distinguish what respondents literally expressed from any higher-level",
    "contextual interpretation."
  ),
  generate = FALSE
)

textual_group <- names(nail_evidence(textual_preview)$groups)[[1L]]
textual_evidence <- nail_evidence(textual_preview, select = textual_group)
textual_evidence$metrics
cat(substr(nail_prompt(textual_preview, select = textual_group, print = FALSE), 1, 1100), "\n")

textual_prep_preview <- nail_textual_prep(
  dataset = fabric_A,
  num.var = 4,
  num.text = 3,
  sample.pct = 1,
  generate = FALSE
)
```

With `generate = TRUE`, inspect the raw response and parsed profile. Do not
present a generated thematic synthesis as a literal statement by respondents.

```{r textual-generation, eval = FALSE}
textual_generated <- nail_textual(
  dataset = fabric_A,
  num.var = 4,
  num.text = 3,
  isolate.groups = TRUE,
  sample.pct = 0.35,
  seed = 123,
  provider = "ollama",
  model = "mistral-small3.2",
  generate = TRUE
)

nail_response(textual_generated, select = textual_group, print = FALSE)
```

## Sensory product space: `nail_qda_space()`

`nail_qda_space()` builds a PCA space from QDA adjusted means and characterizes
its dimensions with latent CONDES evidence. It is therefore a downstream
workflow: the QDA evidence remains the statistical foundation.

```{r qda-space}
qda_space_preview <- nail_qda_space(
  qda_preview,
  ncp = 2,
  expertise_mode = "sensory",
  generate = FALSE
)

qda_axis <- names(nail_evidence(qda_space_preview)$axes)[[1L]]
nail_evidence(qda_space_preview, select = qda_axis)
cat(substr(nail_prompt(qda_space_preview, select = qda_axis, print = FALSE), 1, 1100), "\n")
```

If model-assisted QDA product interpretations are reused in a generated
space interpretation, NaileR emits one non-blocking warning. The axis result
should then be confronted with the original QDA and PCA/CONDES evidence. The
warning is not emitted for a preview, statistical-only evidence, or expert-only
product summaries.

# Composing CATDES and TEXTUAL

`nail_catdes_textual()` combines an existing CATDES result with an existing
TEXTUAL result. The intended direction is:

```text
global CATDES interpretation
        ↓
verbatims illustrate, deepen, qualify, or question the portrait
        ↓
enriched global group interpretation
```

Individual closed-question configurations are optional contextual resources,
not a mandatory second analysis. The CATDES statistical evidence remains the
check on the global portrait, and the text registry preserves the provenance
of quotations.

The historical Atomic Habits example uses the existing group variable and
verbatim column. It requires generated upstream profiles, so it is shown but
not executed while building this vignette.

```{r catdes-textual-generation, eval = FALSE}
data(atomic_habit_clust, package = "NaileR")

atomic_catdes_data <- atomic_habit_clust[, c(1:20, 51), drop = FALSE]
atomic_textual_data <- data.frame(
  group = atomic_habit_clust[[51]],
  text = atomic_habit_clust[[31]],
  stringsAsFactors = FALSE
)

atomic_catdes <- nail_catdes(
  dataset = atomic_catdes_data,
  num.var = ncol(atomic_catdes_data),
  interpretation_mode = "latent",
  isolate.groups = TRUE,
  provider = "ollama",
  model = "mistral-small3.2",
  generate = TRUE
)
atomic_textual <- nail_textual(
  dataset = atomic_textual_data,
  num.var = 1,
  num.text = 2,
  isolate.groups = TRUE,
  sample.pct = 0.40,
  seed = 123,
  provider = "ollama",
  model = "mistral-small3.2",
  generate = TRUE
)

atomic_enriched <- nail_catdes_textual(
  catdes = atomic_catdes,
  textual = atomic_textual,
  isolate.groups = TRUE,
  provider = "ollama",
  model = "mistral-small3.2",
  generate = TRUE
)

nail_evidence(atomic_enriched, select = "1")
nail_prompt(atomic_enriched, select = "1", print = FALSE)
nail_response(atomic_enriched, select = "1", print = FALSE)
```

If you ask the LLM to interpret text based on an already generated cluster
summary, NaileR shows a warning: an error in the first step could carry over
into the second. The generated summary is not primary statistical or textual
evidence, so the enriched result should be checked against the original CATDES
evidence and text registry. This is a reason for review, not evidence that the
composition is automatically invalid.

# Practical checklist

Before relying on an interpretation:

1. Run the analysis with `generate = FALSE`.
2. Inspect the underlying evidence (or statistical results) with `nail_evidence()`.
3. Inspect the exact prompt with `nail_prompt()`.
4. Check that the introduction supplies context rather than hidden evidence.
5. Check that the request asks a scientifically meaningful question.
6. If generating, preserve and inspect the raw response with `nail_response()`.
7. Separate measured facts, statistical synthesis, substantive interpretation,
   and broader hypotheses.
8. For composed workflows, return to the upstream evidence when a warning is
   emitted or when a generated interpretation appears unusually certain.

The package is designed to make this review possible. It does not remove the
need for domain expertise or human control of the final interpretation.

# Historical compatibility and scope

The current evidence-first workflows coexist with historical exports such as
`nail_sort()` and the LLM distance/similarity functions. Those functions are
outside the active stabilization scope of this vignette and are not silently
removed or refactored here. Compatibility paths such as
`nail_textual_contextualized()` remain available for existing users; new
analyses should prefer the evidence-first functions documented above.
